String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-25 2023-1-19/sections/introduction.tex
intro

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{5, 60\}} \wedge x \in \Sigma^* c \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with $c$, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in  both $(\Sigma \setminus a)^{\{5, 60\}}$ and $(\Sigma \setminus b)^{\{5, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c$ requires that $x$ has to end with $c$.

@@

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting into a new version of OSTRICH, called $\ostrichrecl$ for convenience.  
Furthermore, we utilize a collection of benchmark suites, which comprise 49,379 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contains sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experimental results show that $\ostrichrecl$ can solve 49,128 out of 49,379 instances, while the state-of-the-art best solver Z3str3RE solves only 47,407 instances, with a timeout of 60 seconds. Moreover, $\ostrichrecl$ and Z3str3RE spend almost the same amount of time in average, that is, 2.5 seconds, in solving each instance. In comparison, CVC5, Z3str3, and Z3seq solve only 44,001, 36,320, 46,813 instances respectively, with an average time 7.8, 7.7, 3.8 seconds per instance.

>>>>

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with $c$, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in  both $(\Sigma \setminus a)^{\{5, 60\}}$ and $(\Sigma \setminus b)^{\{5, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c$ requires that $x$ has to end with $c$.

@@

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting into a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
% 
Furthermore, we utilize a collection of benchmark suites, which comprise 49,379 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contains sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experimental results show that $\ostrichrecl$ can solve 49,128 out of 49,379 instances, while the state-of-the-art best solver Z3str3RE solves only 47,407 instances, with the timeout set as 60 seconds. Moreover, $\ostrichrecl$ and Z3str3RE spend almost the same amount of time in average, that is, 2.5 seconds, in solving each instance. 
In particular, the string constraint in~(\ref{eqn-running}) can be solved by $\ostrichrecl$ within one second. 
In comparison, CVC5, Z3str3, and Z3seq solve only 44,001, 36,320, 46,813 instances respectively, with an average time 7.8, 7.7, 3.8 seconds per instance.
